On the analytic structure of the linear response susceptibility in open systems

### 2018-03-06 AnalyticStructure_main.tex
update

\documentclass[a4paper]{article}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}

%% Sets page size and margins
\usepackage[a4paper,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

%% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools,braket}
\usepackage{xifthen}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{authblk}

@@

\title{Critical assessment of Linear-Response Excitations in Molecular system\\
or \\
On the Localization of the Linear-Response excitations in Open Systems}
\author[2]{Marco D'Alessandro}
\author[1]{Luigi Genovese}
\affil[1]{Laboratoire de Simulation Atomistique (LSim), SP2M, INAC, CEA-UJF, 17 Av. des Martyrs,
38054 Grenoble, France}
\affil[2]{Istituto di Struttura della Materia-CNR (ISM-CNR), Via del Fosso del Cavaliere 100, 00133 Roma, Italia}

\begin{document}
\maketitle

\begin{abstract}
We analyze how the fundamental equations which govern the linear response properties of a quantum system are influenced by the presence of open boundaries.
In particular we focus our attention on linear-response (LR)-TDDFT calculation of molecular systems. We show that optical excitations of energy below the LR ionization potential behave as observable, localized quantities, that can be computationally calculated with high precision, provided an adequate level of completeness.
Above this energy threshold, the analytic structure of the susceptivity functional might be separated in two sectors: one can still be described in terms of observable excitation with discrete energies, that can be captured within computational setups tailored for localized states. However such excitations are embedded in a continuum of excited states that, on the contrary, behave as plane waves and cannot be described efficiently by the same computational techniques. Such excitations do not exhibit observable features and \emph{cannot} be considered as poles of the dynamical polarizability. Nonetheless, the contribution of such continuum of excitation reveals to be of paramount importance even for expressing LR quantities in the infrared regime like for example static polarizabilities.
This is particularly evident for small molecules; we show how these considerations might be influenced by the size of molecular system.
 We then present indicators that can help to quantify such potential observable property of an excitation, that can be evaluated 
in any discretization scheme.  This result is a inherent behavior of the system's Liouvillian, and does not depend on computational treatment of the 
unoccupied subspace.
\end{abstract}

\section{On the analytic structure linear response susceptibility}
\begin{itemize}
\item People want converged quantities related to systems properties. How to obtain them?
\item FS naturally emerges as the minimal building block for the evaluation of physical quantities in LR formalism . For a given Phi, I need to express O and FS.
\item We imposing a pertubation of the system, it is easay to see that the FS are also  functional of the Perturbation potential applied on the response function of the system. FS = R Phi for the same omega.
\item the fundamental objet is therefore R.
\item[$\Rightarrow$] EoM of FS evidence two distinct regimes of \emph{locality}; 
below threshold FS are localized quantities and can be expressed explicitly;
above threshold they behave as plane waves and the only physical meaning is inside the scalar product, which requires bith  the evaluation of $\op \Phi$ and $\op O$.
\item this fact reflect the physical effect that a low-energy excitation is not able to ionize the molecule.
\item As by hypothesis we are much more capable to express localized quantities it is therefore likely that it easier to converge LR quantities in the infrared regime. Let us show why this fact happens.
\item If I have the correct DoF to express the FS as a genuinely localized quantity I have the LR for \emph{any} observable $\op O$. This fact would not be possible above threshold.
\item The capability of expressing the FS below threshold reflects the capability of expressing the LR kernel. Therefore \emph{the same} localization argument should be reformulated in term of R
\item Let us discuss how this observations are related to the analytic structure of the LR.
\item FS are explicitly dependent on the perturbation potential. The LR would be (formally) completely defined in terms of free oscillation of the system (eigenoperators of the Liouvillian).
\item The FS can be also expressed as a sum over excitations
%\item As the imaginary part of the FS should be also localized below threshold this implies that also the excitations are localized quantities when E is below threshold
\item [$\Rightarrow$]Such sum over the excitation must be such as to express a localized quantity below threshold, (at least when applied to localized potentials).
\item[$\Rightarrow$] In addition, also the excited states undergo two different regimes of locality, but this time their EoM implies \emph{also} conditions on the discreteness of the energies. Above threshold the excitations belong to a \emph{continuum}
\item Such continuum of excitation is formed by a set of ldelocalized excitations that are nontheless important to express the localized infomration below threshold.
\item However this implies that \emph{also} the analytic structure of the LR undergo two different regimes.
%\item The absorption spectra, which are only dependent by the imaginary part of R, are therefore well expressed only for discrete energies
\item[$\Rightarrow$] We might therefore show that below threshold the analytic structure of the LR is given in terms of localized quantities that are at discrete, observable and physically meaningful energies.
\item[$\Rightarrow$] The spectral function of the LR, which would be written as a sum of delta, only converges when the ecxitation are dense enough in the continuum sector, wheras convergence in the infrared regimes only requires identification of the discrete excitation energies
\item There are multiple thresholds, one for any QP energy. The first threshold is the ionization potential.
\item[$\Rightarrow$] in a computational setup that is able to capture localized information, the analytic structure of the LR can be expressed with arbitrary good convergence \textbf{only} below the first threshold, as localized information it is more easily expressed in computational treatments that are initially performed for the Ground state.
\item For this reason it is easier to have convergence in the optical properties for energy sector below the first threshold
\item Above each of the threshold (therefore after the first), there is a continuum of delocalized excited states that have to be expressed if we want to abandon the representation in terms of scalar product between FS and O.
\item Such representation in terms of O, FS and Phi is still the most convenient to have accurate evaluation of the LR above threshold.

\end{itemize}
\section{Reboot with recent considerations}
\subsection{Perturbative approach to finite-size corrections}
Given a approximate eigenvector $\ket{\psi}$ we approximate its energy by $E=\bra{\psi} \op H \ket{\psi}$.
This energy is an approximation of the true eigenvalue. We assume that the wavefunction gradient $\ket{g(E)}\equiv \left(\op H -E\right)\ket{\psi}$ indicates the distance from the solution of the Schroedinger Equation (SE) in a complete basis set.
By imposing the solution of the SE for a better approximation of the wavefunction $\psi + \Delta \psi$
we should obtain, self-consistently:
\begin{align}
\left(E -\op H \right)\ket{\Delta \psi} &= \ket{g(E+ \Delta E)}\;,\\
\Delta E &= 2 \bra{\psi} \op H \ket{\Delta \psi}\;.
\end{align}
From the expression of $\Delta \psi$ we can recalculate another approximation for the wavefunction and, correspondingly, for the eigenvalue.
By definition, the exact solution is reached only when $\ket{g}=0$, which implies that the norm of $\Delta \psi$ must decrease during the perturbative process.

\textbf{Theorem:} Only the eigenstates with negative energy might have a finite-size correction that can be expressed perturbatvely.

\textbf{Proof}: 
If we apply this process to estimate the finite-size correction to a numerical solution of the Schroedinger equation, as the eigenvalue should converge faster than the eigenvector, we conclude that the gradient $g$ becomes nonzero only in regions that  should be located far from the potential region. If this is true, at the next iteration the gradient would become $g + \op H \Delta \psi + \mathcal{O}(\Delta E)$.
Therefore, for having a convergent perturbative process, also the ket $\ket{\Delta \psi}$ solution of the above equation should be localized far from the support of the potential. In other terms, the relation between $g$ and $\Delta \psi$ should be local in space.

The Lippmann-Schwinger relation defines the Dyson series of the resolvent of the Hamiltonian in terms of the Helmholtz Green's function $\op G_0(E)$:
\be 
\op G(E) \equiv (E- \op H)^{-1} = \op G_0(E) + \op G_0(E) \op V \op G(E)\;.
\ee
We have therefore 
\be 
\ket{\Delta \psi} = 
\sum_{p=0}^\infty \op G_0(E) 
\left(\op V \op G_0(E)\right)^p \ket{g}\;.
\ee
For the properties of the Helmholtz equation, locality in space of $\Delta \psi$ might only be recovered for negative energy values. This equation proves that for positive values of the energy the finite-size corrections will \emph{always} provide fully delocalized corrections: it would be impossible to observe a convergence of the wavefunction in the interior region by simply increasing the size of the simulation domain. 
If the value of the approximate energy does not become negative during the perturbative process, 
a modification of the domain boundaries will impact the wavefunction in the entire simulation region. In this case, convergence might only be achieved uniformly and there is no guarantee that a larger domain will provide similar energies.
$\square$
\begin{itemize}
\item Write the excitations as the eigenstates of the Liouvillian:
\begin{itemize}
\item The Casida Matrix provides the eigenvectors in the basis of the unperturbed transitions
\item From the Casida eigenproblem we 
get the poles of the dynamical polarizability and we may reconstruct the perturbed wavefunctions
\item Lot of solution in state of 
the art to express \emph{the same information}, still by avoiding the explicit diagonalization of the problem
\end{itemize}
\item Such transitions might also be written in terms of the solutions of the eigenproblem for the perturbed hamiltonian
\begin{itemize}
\item There are therefore excitations which correspond to localized solutions of the above eigenproblems
\item There is a \emph{energy threshold} above which the eigenfunction associated \emph{cannot be anymore} localized
\end{itemize}
\item Below threshold, the eigenfunction are localized and the energy values are \emph{observable} quantities (there is a convergence in terms of the simulation parameters)
\begin{itemize}
\item Casida treatments for different
basis sets exhibit convergence of the low-energy excitations
\item Still, even for such localized excitations, non-trivial dependence on the empty states on some of excitation value is observed (energies converge relatively slowly wrt $N_\alpha$)
\item The locality of the excitation can be \emph{measured}
{\bf NEW idea for the SoB: why not to consider}
$$
\int \mathrm d \mathbf r
\rho(\mathbf r) |\phi_p^E(\mathbf r)|^2
$$
or similar quantities, that are independent on the number of empty and  bound states
\item Localized Excitation are the  only one with an observable character (this is surely already known)
\item The spectrum of the Liouvillian in this regime is a \emph{discrete} spectrum; the excitation energies are \emph{poles} of the dynamical polarizability
nano\item We may associate the value of the threshold to the \emph{ionization potential}: This is a very interesting remark in my opinion: the physics of the optical excitation is observable 
(and therefore localized) only if the energies are not higher than the energy needed to ionize the system. After this point, it seems normal to me that the ``optical'' excitations with 
real value of the energy are not  meant to be observable anymore.
\item Optical Excitations below threshold have therefore a (very) long lifetime (poles on the real axis), and they contribute to the analytic structure of the dynamical polarizability.
\end{itemize}
\item Above this threshold the energies have a strong dependence on the computational treatment
\begin{itemize}
\item The high-energy part of the 
spectrum does not converge with the 
boundary conditions, even though they converge wrt the empty states (established techniques \emph{do not solve} such problem)
\item This is \emph{not} a problem that is related to the continuum collapse of the empty states.
\item They belong to the continuum of  the Liouvillian $\rightarrow$ they collapse, or depend strongly on the basis set adopted
\item They cannot be considered separately \emph{even when they are expressed in localized basis} as 
their shape is dependent on the treatment
\item They energies are not \emph{true} poles of the dynamical polarizability (the spectra converge only by superposition of deltas)
\end{itemize}
\item Optical excitation above Ionization threshold are therefore observable \emph{only} if they possess a lifetime (the peaks of the spectrum have a well-defined broadening). 
In this way they might be expressed in terms of resonant wavefunctions
\end{itemize}

\section{General theoretical setup}

\subsection{Basic definitions}\lb{BasicDef}

We consider a generic one-body quantum mechanical problem described by the hermitian Hamiltonian $\hnot$  written in terms of the standard 
kinetic operator $\op T$ plus a one-body potential $\op V$, which we assume to be completely general, i.e. possibly nonlocal.
We consider open systems, so that $\hnot$ is defined on $\mathbb L_2(\mathbb R_3)$ and the basis of its eigenfunctions spans this 
functional space. Therefore, together with the discrete (bound) part of its spectrum $\eps_b$, the Hamiltonian may have an essential spectrum 
$\eps(c)$, parametrized by the continuum label $c$.

A generic state $\ket{f} \in \mathbb L_2(\mathbb R_3)$ can be expanded in the basis of the eigenstates of $\hnot$, that is:
\be\lb{ketBasisExpDef1}
\ket{f} = \sum_b f_b \ket{\psi_b} + \int \dd c f(c) \ket{\psi(c)}  = 
\sint \dd s f(s) \ket{\psi(s)} \;,
\ee
where we have introduced the compact multi-index $s$ that collect both the discrete and continuum part of the spectrum of $\hnot$.
The normalization of the basis can be chosen so that the standard orthonormality condition holds, that is: 
\be
\brket{\psi(s)}{\psi(s')} = \delta(s-s') \;,
\ee
and the completeness condition is written as:
\be\lb{completenessH0basis1}
\identity = \sint \dd s \ketbra{\psi(s)}{\psi(s)} = \sum_{b} \ketbra{\psi_b}{\psi_b} + \int \dd c \ketbra{\psi(c)}{\psi(c)} \;. 
\ee
Thanks to this condition the coefficients of the expansion \eqref{ketBasisExpDef1} can be expressed via a scalar products, namely $f_{b} = 
\brket{\psi_b}{f}$ and $f(c) = \brket{\psi(c)}{f}$.

\subsection{Time evolution of states in the linear response regime}

We discuss the time evolution of a generic state vector in the case in which a hermitian time dependent perturbation $\hone{t}$ is added to $\hnot$. 
To achieve this task we have to solve the time dependent Schroedinger equation supplied by the knowledge of the state at the ``initial'' time $t_0$.
We seek for a solution of the form:
\be\lb{ketTimeEvol1}
\ket{\psi(t_{0},t)} = e^{-\ii \hnot t} \op{U}_{I}(t,t_0) e^{\ii \hnot t_0}\ket{\psi_0} \; . 
\ee
where $\op{U}_I(t,t_0)$ is the time evolution operator induced by $\hone{}$ that satisfies the ``initial'' time condition $\op{U}_I(t_0,t_0)=1$. 
Plugging the expression \eqref{ketTimeEvol1} in the Schroedinger equation we obtain a differential equation for $\op{U}_I$ that can be solved 
iteratively as a power series in $\hone{}$. Limiting to the linear order provides:
\be \lb{UiLinOrderSol1}
\op{U}_I(t,t_0) = \identity - \ii \intdef{t_0}{t}{t'} e^{\ii \hnot t'} \hone{t'} e^{-\ii \hnot t'} \; .
\ee
So, at the first order in $\hone{}$ equation \eqref{ketTimeEvol1} can thus be written as:
\be \lb{ketTimeEvol2} 
\ket{\psi(t_{0},t)} = \left( \op{U}_0(t-t_0) - \ii \intdef{t_0}{t}{t'} \op{U}_0(t-t') \hone{t'} 
\op{U}_0(t'-t_0) \right)\ket{\psi_0} \; ,
\ee
were we have defined the time evolution operator of the unperturbed Hamiltonian by
\be
\op{U}_0(t-t_0)\equiv e^{-\ii \hnot(t -t_0)} \;.
\ee
The time evolution of left vectors can be derived in the same way. We can express the temporal evolution of the left 
vectors in the form:
\be \lb{braTimeEvol1}
\bra{\psi(t_{0},t)} = \bra{\psi_0} e^{-\ii \hnot t_0} \op{U}_{I}(t_0,t) e^{\ii \hnot t} \;,
\ee
plugging this equation in the conjugate Schroedinger equation and pursuing the same procedure defined above, we obtain the 
linear order expression of time evolution of left vectors:
\be\lb{leftVectTimeEvol2}
\bra{\psi(t_{0},t)} = \bra{\psi_0} \left( \op{U}_0(t_0-t) + \ii \intdef{t_0}{t}{t'} \op{U}_0(t_0-t') 
\hone{t'} \op{U}_0(t'-t) \right) 
\ee
It is immediate to check that, in order to preserve the norm vectors during the time evolution, it is sufficient that $\op{U}_I$ 
possesses the  semi-group structure $\op{U}_I(t_2,t_1) \op{U}_I(t_1,t_0) = \op{U}_I(t_2,t_0)$. An inspection of \eqref{UiLinOrderSol1}
evidences that this condition is satisfied in the linear response regime.

\subsubsection{Time evolution of stationary states. Adiabatic limit vs finite time switch-on procedure}

We are interested in determining the time evolution of an eigenstate of $\hnot$ under the action of the perturbation $\hone{t}$ in the linear 
response regime. A proper statement of the problem requires the specification of the way in which the perturbation is switched on.
Possible choice of the switching on protocol are, for instance, the \emph{adiabatic limit} and the \emph{sudden limit} protocols. In the first
case the perturbation is switched on at $t_0=-\infty$, but the intrinsic time structure of $\hone{}$ is modified by the damping function 
$e^{-\eta_a |t|}$. In the second one the perturbation is abruptly switched on at the finite time $t_0$, so that the time structure of
$\hone{}$ is multiplied times the $\theta$-function $\theta(t-t_0)$.

Starting from the generic time evolution equation \eqref{ketTimeEvol2} and choosing the initial condition in the form 
$\ket{\psi_0} = e^{-\\i \eps_b t_0}\ket{\psi_b}$, we can write the time evolution of the generic $\ket{\psi_b}$ eigenvector of $\hnot$ as:
\be \lb{ketTimeEvol3} 
\ket{\psi_b(t)} = e^{-\ii \eps_b t}\left( \identity - \ii \intdef{t_0}{t}{t'} \op{U}_0(t-t') \op\Phi^{a.l,s.l}(t') \op{U}_0(t'-t)  \right)\ket{\psi_b} \; ,
\ee
where we have introduced the \emph{perturbing field operators}: 
\be
\op\Phi^{a.l.}(t) = e^{-\eta_a |t|} \hone{t} \;, \qq
\op\Phi^{s.l.}(t) = \theta(t-t_0) \hone{t}
\ee
that codify the effective of the perturbing Hamiltonian in the specific switch-on procedure. Making usage of elementary properties of $\theta$ functions allows us
to rewrite the former time integral as:
\be\lb{AopDef1}
\op A(t) = \intdef{-\infty}{\infty}{t'} \op{U}^{+}_0(t-t') \op\Phi^{a.l,s.l}(t') \op{U}^{-}_0(t'-t) \;,
\ee
so that the effect of the switch-on protocol are completely reabsorbed in the perturbing field. In the subsequent analysis we will omit the apex in the fields $\op\Phi$, 
unless explicitly needed, to keep notations as simple as possible. Thanks to this definition we can express the time evolution of the eigenvectors of $\hnot$ in the 
compact form:
\be\lb{ketBraTimeEvol1}
\ket{\psi_b(t)} = e^{-\ii \eps_b t}\left( \identity - \ii \op A(t)  \right)\ket{\psi_b} \; , \qq 
\bra{\psi_b(t)} = \bra{\psi_b} \left( \identity + \ii \op A(t)  \right) e^{\ii \eps_b t} \; .
\ee

\subsection{Linear response on the density operator. Liouville equation in the time domain}

We introduce the \emph{density operator} defined as the sum of the one-particle densities of all the states with energy lower than the \emph{chemical potential} 
(Fermi energy):
\be\label{rhodef1}
\dmnot = \sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}  \;,
\ee
where the index $p$ labels the occupied state, \emph{i.e} the subset of the bound eigenstates of $\hnot$ with energy lower than the Fermi energy.
Clearly $\dmnot$ commutes with $\hnot$ and in the unperturbed case is a time-independent operator. When the perturbation is switched the system starts to evolve
and we indicate with $\dm(t)$ the modified density of the perturbed system. We introduce the shorthand notation:
\be\lb{rhodef2}
\dm(t)  =  \dmnot + \dm'(t)\;,
\ee
where $\dm'(t)$ is called the \emph{response density operator} and represents the time dependent modification of the ground state density induced by the perturbation.
Here we are assuming that the highest occupied energy level and the lowest unoccupied one will never intersect during the time evolution. Thanks to this condition, the 
linear response regime the expression of $\dm'(t)$ can be readily derived using equations \eqref{ketBraTimeEvol1} and reads:  
\be\lb{rhoTimeEvolutionDef3}
\dm'(t) = \ii \commutator{\dmnot}{\op A(t)} \;.
\ee
The equation of motion (Liouville equation) for $\dm'$ is obtained by performing the time derivative of the previous equation. The derivative of $\op A(t)$ can be 
computed starting from its definition and reads:
\be
\d_t\op A(t) = \op\Phi(t) -\ii \commutator{\hnot}{\op A(t)} \nn \;,
\ee
and using the Jacobi identity we have:
\be\lb{LiouvilleRhoDef1}
\ii \d_t \dm'(t) = \commutator{\op\Phi(t)}{\dmnot} + \ii\commutator{\hnot}{\commutator{\dmnot}{\op A(t)}} = 
\commutator{\op\Phi(t)}{\dmnot} + \commutator{\hnot}{\dm'(t)} \;.
\ee
It is useful to introduce the \emph{Liouvillian} superoperator:
\be\lb{L0sopDef1}
\Liouvnot \equiv \commutator{\hnot}{\cdot} \;,
\ee
so that the Liouville equation can be written as:
\be\lb{LiouvilleRhoDef2}
\ii \d_t \dm'(t) = \Liouvnot \dm'(t) + \commutator{\op\Phi(t)}{\dmnot} \;.
\ee

\subsection{Self-consistent Hamiltonians. Generalized Liouville operator}

Let us extend the the former analysis to self-consistent systems, \emph{i.e.} to the case in which the unperturbed Hamiltonian is itself a function of the density.
In this case the perturbing field $\op\Phi$ has also the effect of adding a further perturbation operator to the perturbing Hamiltonian. In the linear response framework, 
this latter quantity can be expressed as the functional derivatives of $\hnot$ with respect to the density kernel (evaluated at $\dm=\dmnot$) times the value of the
induced density modification, that is:
\be\label{Vptspacedef1}
\op V'(t)= 
\int \dd \r \dd \r' \frac{\delta\hnot}{\delta \rho_0(\r,\r')}
\trace{\dm'(t) \ket{\r} \bra{\r'}} \;.
\ee
The Liouville equation for self-consistent systems can be written by adding the contribution of $\op V'$ to $\op\Phi$ in \eqref{LiouvilleRhoDef2}. 
It reads:
\be\lb{LiouvilleRhoSelfDef1}
\ii \d_t \dm'(t) = \Liouvnot \dm'(t) + \commutator{\op V'(t)}{\dmnot} + \commutator{\op \Phi(t)}{\dmnot} \;.
\ee
We point out that the response density operator obeys a self-consistent equation, since the  induced perturbing field $\op V'$ is itself a function of $\dm'$. 
It is possible to formally recast \eqref{LiouvilleRhoSelfDef1} in the same form of \eqref{LiouvilleRhoDef2} by introducing the \emph{coupling}
superoperator:
\be\lb{couplingSopDef1}
\coupl = 
\int \dd \r \dd \r' \commutator{\frac{\delta\hnot}{\delta \rho_0(\r,\r')}}{\dmnot}
\trace{\cdot \ket{\r} \bra{\r'}} \;,
\ee
in this way the first commutator in \eqref{LiouvilleRhoSelfDef1} can be written as the formal action of the coupling superoperator to $\dm'$ and \eqref{LiouvilleRhoSelfDef1}
can be written as:
\be\lb{LiouvilleRhoSelfDef2}
\ii \d_t \dm'(t) = \Liouv \dm'(t) + \commutator{\op \Phi(t)}{\dmnot} \;, \qq
\Liouv = \Liouvnot + \coupl \;.
\ee

\subsection{Formulation in the frequency domain}

We reformulate the Liouville equation \eqref{LiouvilleRhoSelfDef2} in the frequency domain. In order to achieve this task it is
preferable to directly compute the action of the Liouville operator on the Fourier transform of the response density, since in this
way some subtleties that concerns the analytic structure in the $\omega$ complex plane can be properly taken into account.
We first define the response density in the frequency domain that, staring from \eqref{rhoTimeEvolutionDef3}, is given by:
\be\lb{rhoPrimeOmegaDef1}
\dm'(\omega) \equiv \intover{t} e^{\ii \omega t} \dm'(t)
= \ii \commutator{\dmnot}{\op A(\omega)} \;,
\ee
where $\op A(\omega)$ is The Fourier transform of \eqref{AopDef1} and reads:
\be \lb{AomegaDef1}
\op A(\omega) = \frac{1}{2\pi} \int_{-\infty}^\infty  \dd \omega'
\op G_0^+(\omega') \op \Phi(\omega) \op G_0^-(\omega'-\omega)\;.
\ee
The action of $\Liouvnot$ on $\dm'(\omega)$ follows from \eqref{L0sopDef1}:
\be\lb{LiouvillianRhopomegaDef1}
\Liouvnot \dm'(\omega) = \ii \commutator{\hnot}{\commutator{\dmnot}{\op A(\omega)}} = 
\ii \commutator{\dmnot}{\commutator{\hnot}{\op A(\omega)}} \;,
\ee
and the commutator between the unperturbed Hamiltonian and $\op A(\omega)$ can be directly computed starting from \eqref{AomegaDef1}: 
\be
\commutator{\hnot }{\op A(\omega)} = \bbomega \op A(\omega) - \ii \op \Phi(\omega)\;, \nn
\ee
where $\bbomega \equiv \omega + 2\ii\eta$, so that:
\be\lb{LiouvillianRhopomegaDef2}
\Liouvnot \dm'(\omega) =  \bbomega\dm'(\omega) + \commutator{\dmnot}{\op\Phi(\omega)}\;.
\ee
The coupling superoperator has the same formal action on the response density both in the time and in the frequency domain, so the complete
Liouville equation in $\omega$ space can be written as:
\be\lb{LiouvillianRhopomegaDef3}
\left(\bbomega - \Liouv\right) \dm'(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot}\;.
\ee

\section{Critical Analysis of the linear response results}

The main aim of this section is to discuss the reliability in the assessment of the results provided by the linear response formalism. In particular we are interested in 
analyzing the \emph{linear response functional} defined as:
\be\lb{LinearResponseFunctDef1}
{\cal R}_{\op O}(\omega) = \trace{\dm'(\omega)\op O} \;,
\ee
that represents the perturbation-induced modification of the expectation value of an observable $\op O$. Here, for a sake of concreteness, we are limiting our considerations 
to observables that do not explicitly depend on time.

We start this analysis by deriving an ulterior expression of the response density operator in the frequency space. We consider equation \eqref{rhoPrimeOmegaDef1} and observe that, for any 
operator $\op O$, we may write:  
\be
\commutator{\dmnot}{\op O} = \dmnot \op O \op Q_0 - \op Q_0 \op O \dmnot\;,
\ee
where $\op Q_0= \identity - \dmnot$ is the projection operator in the subspace orthogonal to $\dmnot$. So, thanks to this relation we have that:
\be
\dm'(\omega) = \ii\sum_p\left(\ketbra{\psi_p}{\psi_p}\op A(\omega)\op Q_0 - \op Q_0\op A(\omega)\ketbra{\psi_p}{\psi_p}\right) \;.
\ee
This equation suggests us to define the \emph{fluctuation states}:
\be\lb{fluctuationStateDef1}
\ket{f_p(\omega)} \doteq -\ii\op Q_0\op A(\omega)\ket{\psi_p} \;, \qq
\bra{f_p(\omega)} \doteq \bra{\psi_p}\ii\op A(-\omega)\op Q_0 \;,
\ee
where the $\omega$ dependence of the bra is consistent with hermitian conjugation property of the Fourier transform of a hermitian operator. An inspection of equation \eqref{ketBraTimeEvol1} 
evidence that formulas \eqref{fluctuationStateDef1} are equivalent to the action of the $\op Q_0$ projector acting on the time evolution of the (occupied) eigenstates of $\hnot$, written 
in the $\omega$ domain. Fluctuation states are, by definition, orthogonal to the occupied eigenstates of $\hnot$, so that:
\be
\brket{f_p(\omega)}{\psi_q} = \brket{\psi_q}{f_p(\omega)} = 0 \;, \forall \, p,q,\omega \;.
\ee
The response density written in terms of the fluctuation states reads:
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;,
\ee
and consequently the linear response functional can be expressed as:
\be\lb{LinearResponseFunctDef2}
{\cal R}_{\op O}(\omega) = 
\sum_p\left(\bra{f_p(-\omega)}\op O\ket{\psi_p} + \bra{\psi_p}\op O\ket{f_p(\omega)}\right) \;.
\ee

>>>>

\documentclass[reprint,aps,prb]{revtex4-1}

%% Language and font encodings
\usepackage[english]{babel}
%\usepackage[utf8x]{inputenc}
%\usepackage[T1]{fontenc}

%% Useful packages
\usepackage{graphicx}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools,braket}
\usepackage{xifthen}

@@

\begin{document}

%\preprint{APS/123-QED}

\title{On the analytic structure of the linear response susceptibility in open systems}

\author{Luigi Genovese}
\affiliation{Laboratoire de Simulation Atomistique (LSim), SP2M, INAC, CEA-UJF, 17 Av. des Martyrs,
38054 Grenoble, France}
\author{Marco D'Alessandro}%
\affiliation{Istituto di Struttura della Materia-CNR (ISM-CNR), Via del Fosso del Cavaliere 100, 00133 Roma, Italia}

\date{\today}

\begin{abstract}
By performing linear-response (LR)-TDDFT calculation of molecular systems within a highly complete real-space basis set, we analyze how the behavior of the dynamical 
polarizability is influenced by the discretization conditions of the computational treatment.
We show that optical excitations of energy below the LR ionization potential behave as observable, localized quantities, that can be studied with high precision, 
provided an adequate level of completeness. We then present indicators that can help to quantify such potential observable property of an excitation, that can be evaluated 
in any discretization scheme. Under this light, we also show that excitation energies above ionization threshold do not exhibit such observable features and \emph{cannot} be 
considered as poles of the dynamical polarizability. This result is a inherent behavior of the system's Liouvillian, and does not depend on computational treatment of the 
unoccupied subspace.
\end{abstract}

%\pacs{Valid PACS appear here}
\maketitle

\section{Introduction}

.....

\section{Theory}

In the framework of the linear response formalism one is typically interested in computing the \emph{linear response functional}:
\be\lb{LinearResponseFunctDef1}
{\cal R}_{\op O}(\omega) = \trace{\dm'(\omega)\op O} \;,
\ee
that represents the perturbation-induced modification of the expectation value of an observable $\op O$\footnote{For a sake of concreteness, we are limiting our considerations 
to observables that do not explicitly depend on time}. Here $\dm'(\omega)$ is the \emph{response density operator} that describes the $\omega$-dependent modification of the density 
operator induced by the perturbation.

\subsection{Basic equation for the linear response formalism for electronic excitations}

The linear response formalism can be effectively applied to the analysis of the electronic excitations of molecular and condensed-matter systems, described by the (density-dependent) 
ground-state hermitian Hamiltonian $\hnot$, coupled to an external perturbation described by the perturbing field $\op\Phi(\omega)$. The $\omega$ structure of $\op\Phi(\omega)$ codifies both 
the intrinsic time dependence of the perturbation and the switching-on procedure, \emph{e.g.} adiabatic or sudden limit protocols.

The response density is expressed as the first-order variation of ground state density
\be\label{rhodef1}
\dmnot = \sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}  \;,
\ee
where $\{\psi_p\}$ denote the occupied states, \emph{i.e} the subset of the bound eigenstates of $\hnot$ with energy lower than the Fermi level.
By considering the linear order time evolution of these states in presence of the perturbation we can express the response density in the $\omega$ domain as follows:
\be\lb{rhoPrimeOmegaDef1}
\dm'(\omega) = \ii \commutator{\dmnot}{\op A(\omega)} \;,
\ee
where $\op A(\omega)$ is a sum of two terms, both written as a triple convolution with the retarded/advanced Green functions $\op G_0^{\pm}$ of the unperturbed Hamiltonian:
\be
\op A(\omega) = \frac{1}{2\pi} \int_{-\infty}^\infty  \dd \omega'
\op G_0^+(\omega') \left(\op \Phi(\omega)+\op V'(\omega)\right) \op G_0^-(\omega'-\omega)\;. \nn
\ee
The presence of the second addend is due to the density dependence of the unperturbed Hamiltonian and describes the perturbation induced by the variation of $\hnot$ under the
action of the perturbing field. In the linear response regime this quantity can be expressed as the functional derivatives of $\hnot$ with respect to the density kernel 
(evaluated at $\dm=\dmnot$) times the value of the induced density modification, that is:
\be\label{VpDef1}
\op V'(\omega)= 
\int \dd \r \dd \r' \frac{\delta\hnot}{\delta \rho_0(\r,\r')}
\trace{\dm'(\omega) \ket{\r} \bra{\r'}} \;.
\ee

\subsection{Fluctuation states. A localization argument for the linear response functional}

We start this analysis by deriving an ulterior expression of the response density operator in the frequency space. We consider equation \eqref{rhoPrimeOmegaDef1} and observe that, for any 
operator $\op O$, we may write:  
\be
\commutator{\dmnot}{\op O} = \dmnot \op O \op Q_0 - \op Q_0 \op O \dmnot\;,
\ee
where $\op Q_0= \identity - \dmnot$ is the projection operator in the subspace orthogonal to $\dmnot$. So, thanks to this relation we have that:
\be
\dm'(\omega) = \ii\sum_p\left(\ketbra{\psi_p}{\psi_p}\op A(\omega)\op Q_0 - \op Q_0\op A(\omega)\ketbra{\psi_p}{\psi_p}\right) \;.
\ee
This equation suggests us to define the \emph{fluctuation states}:
\be\lb{fluctuationStateDef1}
\ket{f_p(\omega)} \doteq -\ii\op Q_0\op A(\omega)\ket{\psi_p} \;, \quad
\bra{f_p(\omega)} \doteq \bra{\psi_p}\ii\op A(-\omega)\op Q_0 \;,
\ee
where the $\omega$ dependence of the bra is consistent with hermitian conjugation property of the Fourier transform of a hermitian operator. Fluctuation states are, by 
definition, orthogonal to the occupied eigenstates of $\hnot$, so that:
\be
\brket{f_p(\omega)}{\psi_q} = \brket{\psi_q}{f_p(\omega)} = 0 \;, \forall \, p,q,\omega \;.
\ee
The response density written in terms of the fluctuation states reads:
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;,
\ee
and consequently the linear response functional can be expressed as:
\be\lb{LinearResponseFunctDef2}
{\cal R}_{\op O}(\omega) = 
\sum_p\left(\bra{f_p(-\omega)}\op O\ket{\psi_p} + \bra{\psi_p}\op O\ket{f_p(\omega)}\right) \;.
\ee

\bibliography{Analytic_biblio}
